Inferring the gravitational potential of the Milky Way with a few precisely measured stars

### 2014-03-15 text/paper2/ms.tex
hide pshock stuff

To model the true, 6D, present-day position of the star, $\W$, we first transform to cartesian, Galactocentric coordinates where ($\bs{r},\bs{v}$) and ($\bs{r}_\sat,\bs{v}_\sat$) are the position and velocity of the star and progenitor. In cartesian coordinates, the present-day position of the star is taken as having been sampled at $t=\tub$ from an isotropic Gaussian centered on one of the instantaneous, effective Lagrange points in position, and an isotropic Gaussian centered on the origin in velocity, and evolved to present day in the parent potential, $\Phi$. We assume that once the star becomes ``unbound'' from the satellite the potential of the satellite can be ignored, and thus treat the star as a test particle.\footnote{This assumption breaks down for sufficiently large-mass progenitors, somewhere between $\sim10^8-10^9~$M$_\odot$.} The satellite mass enters the model through the tidal radius; a changing satellite mass will skew the positions of the effective Lagrange points and velocity scale. We allow for mass-loss by incorporating the initial satellite mass, $m$, and a constant mass-loss term, $\dot{m}$ into the model. Finally, we add two additional parameters: a constant, global offset, $\Loffset$, that scales the position of the effective Lagrange points relative to the classical tidal radius, and a binary parameter $\tailbit$, that is either $-1$ or $+1$ depending on whether the star is in the leading or trailing tail.
\begin{align}
	p(\W \given m, \dot{m}, \W_\sat, \tub, \Phi, \Loffset, \tailbit) &= p(\bs{r} \given \bs{r}_\sat, m, \dot{m}, \tub, \Phi, \Loffset, \tailbit)\,p(\bs{v} \given \bs{v}_\sat, m, \dot{m}, \tub, \Phi)\,\left\vert\J\right\vert_{t=\tub}\\
	p(\bs{r} \given \bs{r}_\sat, m, \dot{m}, \tub, \Phi, \Loffset, \tailbit) &= \mathcal{N}(\bs{r} \given \bs{r}_\sat - \Loffset\,\tailbit\,\rtide\,\hat{x}_1, \sigma^2_r)_{t=\tub}\\
	p(\bs{v} \given \bs{v}_\sat, m, \dot{m}, \tub, \Phi) &= \mathcal{N}(\bs{v} \given \bs{v}_\sat, \sigma^2_v)_{t=\tub}
\end{align}
where $\rtide$ is given by Eq.~\ref{eq:tidalradius}, and $\sigma_v$ by Eq.~\ref{eq:velscale}. $\left\vert\J\right\vert$ is the absolute value of the determinant of the Jacobian that defines the transformation from heliocentric, spherical to Galactocentric, cartesian coordinates. For a single star, in detail, the model works as follows:
\begin{enumerate}
	\item sample from the observed distributions in heliocentric coordinates (Eq.~\ref{eq:obsstar} and Eq.~\ref{eq:obsprog}) to get present-day positions for both the star and progenitor;
	\item transform to Galactocentric, cartesian coordinates, and integrate both orbits backwards (as test particles) for the total interaction time, $\tint$;
	\item evaluate the likelihood at $t=\tub$.
\end{enumerate}

This model works well when all stars are lost due to tidal pruning. Even on a mildly eccentric orbit, stars become preferentially unbound around the satellite's pericenter because of tidal shocking (cite?). We will explore properly modeling this process in future work. For this article, we modify the above to be a mixture model: we add a new parameter for each star, $\pshock$, which is the probability that the star is drawn from a broader, ``shocked'' distribution instead of a tight Gaussian centered on an effective Lagrange point:
\begin{multline}
	p_{\rm strip} = [ \mathcal{N}(\bs{r} \given \bs{r}_\sat - \Loffset\,\tailbit\,\rtide\,\hat{x}_1, \sigma^2_r)\\
		\times \mathcal{N}(\bs{v} \given \bs{v}_\sat, \sigma^2_v) \left\vert\J\right\vert ]_{t=\tub}
\end{multline}
\begin{align}
	p_{\rm shock} = \left [ \mathcal{N}(\bs{r} \given \bs{r}_\sat, 10\sigma^2_r)\mathcal{N}(\bs{v} \given \bs{v}_\sat, 10\sigma^2_v) \left\vert\J\right\vert\right]_{t=\tub}
\end{align}
\begin{multline}
	p(\W \given \W_\sat, \tub, \Phi, \Loffset, \tailbit, \pshock) = \\
		\pshock p_{\rm shock} + (1-\pshock)p_{\rm strip}.
\end{multline}
Here we assume that the ``shocked'' distribution has a much larger variance, and arbitrarily set this to 10 times the variance of the stripped debris. The posterior probability for the position of a star, progenitor, and properties of the parent potential, given the data, is then
\begin{multline}
	p(\Phi, \W, \W_\sat, \tub, \Loffset, \tailbit, \pshock \given \D, \D_\sat) = \\
		\shoveleft{\hspace{3em}\frac{1}{\mathcal{Z}} \, p(\D \given \W) \, p(\D_\sat \given \W_\sat)}\\
			\times p(\W \given \W_\sat, \tub, \Phi, \Loffset, \tailbit, \pshock) \, 
			p(\Phi) \, p(\tub) \, p(\Loffset)
\end{multline}
where the factor $\mathcal{Z}$ only depends on the properties of the data (the \emph{evidence}), and is thus constant when varying the model parameters. All model parameters are summarized in Table~\ref{tbl:params}. Assuming each star is an independent tracer, the full posterior for many stars is then just the product of the individual likelihoods with the priors:
\begin{multline}
	p(\Phi, \W, \W_\sat, \tub, \Loffset, \tailbit, \pshock \given \D, \D_\sat) = \\
		\shoveleft{\hspace{1em}\frac{1}{\mathcal{Z}} \, \prod_i \, p(\D^{(i)} \given \W^{(i)}) \, p(\D_\sat \given \W_\sat)}\\
			\times p(\W^{(i)} \given \W_\sat, \tub^{(i)}, \Phi, \Loffset, \tailbit^{(i)}, \pshock^{(i)}) \, 
			\, p(\tub^{(i)}) p(\Phi) \, p(\Loffset)
\end{multline}

@@

We first test the model using eight particles (stars) --- four from the leading trail, and four from the trailing tail --- randomly sampled from the $2.5\times10^8~M_\odot$ simulation (Section~\ref{sec:sims}), assuming perfect observational data for both the stars and the progenitor. The stars were required to have been stripped after the first pericentric passage and have a present-day distance within $50$ kpc of the Sun. Figure~\ref{fig:} (left panel) shows a sample of stream particles (black) and the eight randomly chosen stars (blue) in the Galactic X-Z plane.

>>>>

To model the true, 6D, present-day position of the star, $\W$, we first transform to cartesian, Galactocentric coordinates where ($\bs{r},\bs{v}$) and ($\bs{r}_\sat,\bs{v}_\sat$) are the position and velocity of the star and progenitor. In cartesian coordinates, the present-day position of the star is taken as having been sampled at $t=\tub$ from an isotropic Gaussian centered on one of the instantaneous, effective Lagrange points in position, and an isotropic Gaussian centered on the origin in velocity, and evolved to present day in the parent potential, $\Phi$. We assume that once the star becomes ``unbound'' from the satellite the potential of the satellite can be ignored, and thus treat the star as a test particle.\footnote{This assumption breaks down for sufficiently large-mass progenitors, somewhere between $\sim10^8-10^9~$M$_\odot$.} The satellite mass enters the model through the tidal radius; a changing satellite mass will skew the positions of the effective Lagrange points and velocity scale. We allow for mass-loss by incorporating the initial satellite mass, $m$, and a constant mass-loss term, $\dot{m}$ into the model. Finally, we add two additional parameters: a constant, global offset, $\Loffset$, that scales the position of the effective Lagrange points relative to the classical tidal radius, and a binary parameter $\tailbit$, that is either $-1$ or $+1$ depending on whether the star is in the leading or trailing tail. The likelihood is then:
\begin{align}
	p(\W \given m, \dot{m}, \W_\sat, \tub, \Phi, \Loffset, \tailbit) &= p(\bs{r} \given \bs{r}_\sat, m, \dot{m}, \tub, \Phi, \Loffset, \tailbit)\,p(\bs{v} \given \bs{v}_\sat, m, \dot{m}, \tub, \Phi)\,\left\vert\J\right\vert_{t=\tub}\\
	p(\bs{r} \given \bs{r}_\sat, m, \dot{m}, \tub, \Phi, \Loffset, \tailbit) &= \mathcal{N}(\bs{r} \given \bs{r}_\sat - \Loffset\,\tailbit\,\rtide\,\hat{x}_1, \sigma^2_r)_{t=\tub}\\
	p(\bs{v} \given \bs{v}_\sat, m, \dot{m}, \tub, \Phi) &= \mathcal{N}(\bs{v} \given \bs{v}_\sat, \sigma^2_v)_{t=\tub}
\end{align}
where $\rtide$ is given by Eq.~\ref{eq:tidalradius}, and $\sigma_v$ by Eq.~\ref{eq:velscale}. $\left\vert\J\right\vert$ is the absolute value of the determinant of the Jacobian that defines the transformation from heliocentric, spherical to Galactocentric, cartesian coordinates. For a single star, in detail, the model works as follows:
\begin{enumerate}
	\item sample from the observed distributions in heliocentric coordinates (Eq.~\ref{eq:obsstar} and Eq.~\ref{eq:obsprog}) to get present-day positions for both the star and progenitor;
	\item transform to Galactocentric, cartesian coordinates, and integrate both orbits backwards (as test particles) for the total interaction time, $\tint$;
	\item evaluate the likelihood at $t=\tub$.
\end{enumerate}
All model parameters are summarized in Table~\ref{tbl:params}. Assuming each star is an independent tracer, the full posterior for many stars is then just the product of the individual likelihoods with the priors:
\begin{multline}
	p(\Phi, \W, \W_\sat, \tub, \Loffset, \tailbit, \given \D, \D_\sat) = \\
		\shoveleft{\hspace{1em}\frac{1}{\mathcal{Z}} \, \prod_i \, p(\D^{(i)} \given \W^{(i)}) \, p(\D_\sat \given \W_\sat)}\\
			\times p(\W^{(i)} \given \W_\sat, \tub^{(i)}, \Phi, \Loffset, \tailbit^{(i)}) \, 
			\, p(\tub^{(i)}) p(\Phi) \, p(\Loffset)
\end{multline}
This model works well when all stars are lost due to tidal pruning. Even on a mildly eccentric orbit, stars become preferentially unbound around the satellite's pericenter because of tidal shocking (cite?). We will explore properly modeling this process in future work.

%For this article, we modify the above to be a mixture model: we add a new parameter for each star, $\pshock$, which is the probability that the star is drawn from a broader, ``shocked'' distribution instead of a tight Gaussian centered on an effective Lagrange point:
%\begin{multline}
%	p_{\rm strip} = [ \mathcal{N}(\bs{r} \given \bs{r}_\sat - \Loffset\,\tailbit\,\rtide\,\hat{x}_1, \sigma^2_r)\\
%		\times \mathcal{N}(\bs{v} \given \bs{v}_\sat, \sigma^2_v) \left\vert\J\right\vert ]_{t=\tub}
%\end{multline}
%\begin{align}
%	p_{\rm shock} = \left [ \mathcal{N}(\bs{r} \given \bs{r}_\sat, 10\sigma^2_r)\mathcal{N}(\bs{v} \given \bs{v}_\sat, 10\sigma^2_v) \left\vert\J\right\vert\right]_{t=\tub}
%\end{align}
%\begin{multline}
%	p(\W \given \W_\sat, \tub, \Phi, \Loffset, \tailbit, \pshock) = \\
%		\pshock p_{\rm shock} + (1-\pshock)p_{\rm strip}.
%\end{multline}
%Here we assume that the ``shocked'' distribution has a much larger variance, and arbitrarily set this to 10 times the variance of the stripped debris. The posterior probability for the position of a star, progenitor, and properties of the parent potential, given the data, is then
%\begin{multline}
%	p(\Phi, \W, \W_\sat, \tub, \Loffset, \tailbit, \pshock \given \D, \D_\sat) = \\
%		\shoveleft{\hspace{3em}\frac{1}{\mathcal{Z}} \, p(\D \given \W) \, p(\D_\sat \given \W_\sat)}\\
%			\times p(\W \given \W_\sat, \tub, \Phi, \Loffset, \tailbit, \pshock) \, 
%			p(\Phi) \, p(\tub) \, p(\Loffset)
%\end{multline}
%where the factor $\mathcal{Z}$ only depends on the properties of the data (the \emph{evidence}), and is thus constant when varying the model parameters. All model parameters are summarized in Table~\ref{tbl:params}. Assuming each star is an independent tracer, the full posterior for many stars is then just the product of the individual likelihoods with the priors:
%\begin{multline}
%	p(\Phi, \W, \W_\sat, \tub, \Loffset, \tailbit, \pshock \given \D, \D_\sat) = \\
%		\shoveleft{\hspace{1em}\frac{1}{\mathcal{Z}} \, \prod_i \, p(\D^{(i)} \given \W^{(i)}) \, p(\D_\sat \given \W_\sat)}\\
%			\times p(\W^{(i)} \given \W_\sat, \tub^{(i)}, \Phi, \Loffset, \tailbit^{(i)}, \pshock^{(i)}) \, 
%			\, p(\tub^{(i)}) p(\Phi) \, p(\Loffset)
%\end{multline}

@@

We first test the model using eight particles (stars) --- four from the leading tail, and four from the trailing tail --- randomly sampled from the $2.5\times10^8~M_\odot$ simulation (Section~\ref{sec:sims}), assuming perfect observational data for both the stars and the progenitor. The stars were required to have been stripped after the first pericentric passage and have a present-day distance within $50$ kpc of the Sun. Figure~\ref{fig:} (left panel) shows a sample of stream particles (black) and the eight randomly chosen stars (blue) in the Galactic X-Z plane.
